### 2014-09-01 thesis-proposal/problem.tex
small revisions to proposal

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
\begin{quote}\textit{The recent development of programming languages suggests that the simultaneous achievement of simplicity 
and generality in language design is a serious unsolved 
problem.}
-- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
%%One might exp ``generality''
%Designing a general-purpose programming language along with its supporting tools (collectively, a \emph{programming system}) has long stood as a grand challenge in computing. 
Achieving both simplicity and generality in a programming language and its associated tools -- collectively, a \emph{programming system} -- requires organizing around a small number of mechanisms that support the realization of many others   as libraries. % In such a system, new dialects and revisions of the system would rarely be necessary. %Indeed, a preponderence of new dialects is clear evidence against claims of generality.  %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
Much progress has been made in the decades since Reynolds' statement above, but a programming system that comprehensively achieves these goals remains elusive, as evidenced by the fact that across all language lineages,  new language and tool {dialects} remain common vehicles for new features and abstractions, both in research and practice.

We observe that these new dialects are often motivated by the desire for new concrete syntax, new type system fragments, or new editor services, all {features} of programming systems that library providers usually do not have substantial control over. %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}. Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally. 
Introducing mechanisms that decentralize control over these features could increase generality, but they must constructed with care, because giving away too much control could weaken the metatheory of the system. Moreover, they must come equipped with \emph{modular reasoning principles} for it to be practical for library clients to rely upon them. Having to manually resolve conflicts between libraries or reestablish important properties of a program any time a new library is imported introduces undesirable complexity. %It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}). For programmers, 
%It can make it difficult to understand and reason about the type of an unfamiliar term, a critical facet of program comprehension (the \emph{typing discipline problem}).

@@

The most common strategy is to ask clients to introduce patterns as strings, deferring their  parsing, compilation and use to run-time. This provides only a partial approximation to feature 1 due to syntactic clashes between string escape sequences and regular expressions. Parsing errors due to these clashes have been observed to be difficult for even experienced programmers to diagnose \cite{Omar:2012:ACC:2337223.2337324}. Moreover, none of the static guarantees can be provided, leading to run-time exceptions (common even in well-tested code \cite{spishak2012type}), and mistakes that are not caught even at run-time, some of which can lead to security vulnerabilities due to injection attacks \cite{owasp2013}. It also introduces performance overhead due to run-time parsing and compilation of patterns, and redundant run-time checks.  Staging these computations can shift some of the errors and cost to compile-time, albeit with some additional semantic and syntactic complexity, but the issues related to the fact that a regular expression is not a string (e.g. syntactic clashes and injection attacks) persist.% parsing can move some errors to run-time, albeit at additional syntactic cost, but it also does not straightforwardly handle splicing-related issues, as we will discuss later.

A more semantically justifiable approach is to dispense with string representations entirely and directly work with an encoding of regular expression patterns using  general-purpose constructs like sums and  products (e.g. as exposed by functional datatypes or objects), or using an abstract type. This provides feature 2a while sacrificing feature 1 (such an encoding affords only an approximation of the verbose abstract syntax of patterns, not its concrete syntax). Advanced invariants like those described in features 2b and 2c are not directly tracked by such encodings and require solutions beyond those commonly found in modern simply-typed languages (we will return to existing approaches  in Sec. \ref{att}).%Explicitly constructing a regular expression by applying datatype constructors is unacceptably verbose.

None of these approaches address the important issue of tooling (feature 3). A number of tools are available online that provide services ranging from simple syntax highlighting to interactive testing and explanations of complex patterns \cite{regexr} and pattern extraction from examples (e.g. \cite{_txt2re:_????}). We have found that programmers benefit substantially from their use \cite{Omar:2012:ACC:2337223.2337324}. Unfortunately, evidence suggests that tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}. We have found that they are also less usable than editor-integrated tools %(e.g. \cite{IntelliJRegexp}) 
because they require the programmer to switch contexts and cannot make use of the code context \cite{Omar:2012:ACC:2337223.2337324}. Working with high-level abstractions like these in a ``low-level'' editor is thus awkward and error-prone, just like attempting to work with high-level abstractions directly in terms of an encoding. For these reasons,   we will consider editor services as within the scope of an abstraction specification.
\subsection{Language-External Approaches}\label{external-approaches}
\begin{figure}
\begin{center}
\includegraphics[scale=.48]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools. (b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active libraries''.}
\label{approaches}
%\vspace{-10px}
\end{figure}

>>>>

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
\begin{quote}\textit{The recent development of programming languages suggests that the simultaneous achievement of simplicity 
and generality in language design is a serious unsolved 
problem.}
-- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
%%One might exp ``generality''
%Designing a general-purpose programming language along with its supporting tools (collectively, a \emph{programming system}) has long stood as a grand challenge in computing. 
Achieving both simplicity and generality in a programming language and its associated tools -- collectively, a \emph{programming system} -- requires organizing around a small number of mechanisms that support the realization of many others   as libraries. % In such a system, new dialects and revisions of the system would rarely be necessary. %Indeed, a preponderence of new dialects is clear evidence against claims of generality.  %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
Much progress has been made in the decades since Reynolds' statement above, but a programming system that comprehensively achieves these goals remains elusive, as evidenced by the fact that across all language lineages,  new language and tool {dialects} remain common vehicles for new features and abstractions, both in research and practice. %Library implementations are not yet satisfying.

We observe that new system dialects are often motivated by a desire for new concrete syntax, new type system fragments, or new editor services, all {features} of programming systems that library providers usually do not have substantial control over. %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}. Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally. 
Introducing mechanisms that decentralize control over these features could decrease the need for new dialects and thus increase generality. But this must done with care, because giving away too much control could weaken important metatheoretic guarantees. Moreover, such  mechanisms must come equipped with \emph{modular reasoning principles} for it to be practical for library clients to rely upon them. Having to manually resolve conflicts between libraries or reestablish important properties of a program whenever a new library is imported introduces undesirable complexity. %It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}). For programmers, 
%It can make it difficult to understand and reason about the type of an unfamiliar term, a critical facet of program comprehension (the \emph{typing discipline problem}).

@@

The most common strategy is to ask clients to introduce patterns as strings, deferring their  parsing, compilation and use to run-time. This provides only a partial approximation to feature 1 due to clashes between string escape sequences and regular expressions (parsing errors due to these clashes have been observed to be difficult for even experienced programmers to diagnose \cite{Omar:2012:ACC:2337223.2337324}). Moreover, none of the static guarantees can be provided, leading to run-time exceptions (common even in well-tested code \cite{spishak2012type}), and mistakes that are not caught even at run-time, some of which can lead to security vulnerabilities due to injection attacks \cite{owasp2013}. It also introduces performance overhead due to run-time parsing and compilation of patterns, and redundant run-time checks.  Staging these computations can shift some of the errors and cost to compile-time, albeit with some additional semantic and syntactic complexity, but the issues related to the fact that a regular expression is not a string (e.g. syntactic clashes and injection attacks) persist.% parsing can move some errors to run-time, albeit at additional syntactic cost, but it also does not straightforwardly handle splicing-related issues, as we will discuss later.

A more semantically justifiable approach is to dispense with string representations entirely and directly work with an encoding of regular expression patterns using  general-purpose constructs like sums and  products (e.g. as exposed by functional datatypes or objects), or using an abstract type. This provides feature 2a while sacrificing feature 1 (such an encoding affords only an approximation of the verbose abstract syntax of patterns, not their concrete syntax). Advanced invariants like those described in features 2b and 2c are not directly tracked by such encodings and require solutions beyond those commonly found in modern simply-typed languages (we will return to this  in Sec. \ref{att}).%Explicitly constructing a regular expression by applying datatype constructors is unacceptably verbose.

None of these approaches address the important issue of tool support (feature 3). A number of tools are available online that provide services ranging from simple syntax highlighting to interactive testing and explanations of complex patterns \cite{regexr} and pattern extraction from examples (e.g. \cite{_txt2re:_????}). We have found that programmers benefit substantially from their use \cite{Omar:2012:ACC:2337223.2337324}. Unfortunately, evidence suggests that tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}. We have found that they are also less usable than editor-integrated tools %(e.g. \cite{IntelliJRegexp}) 
because they require the programmer to switch contexts and cannot make use of the code context \cite{Omar:2012:ACC:2337223.2337324}. Working with high-level abstractions like these in a ``low-level'' editor is thus awkward and error-prone, just like attempting to work with high-level abstractions directly in terms of an encoding. For these reasons,   we see editor services as within the scope of an abstraction specification.
\subsection{Language-External Approaches}\label{external-approaches}
\begin{figure}
\begin{center}
\includegraphics[scale=.48]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools. (b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active libraries''.}
\label{approaches}
%\vspace{-10px}
\end{figure}
